Exploratory data structure comparisons: Three new tools based on Principal Component Analysis

### 2017-03-19 article/article_v0.9.tex
Redid section 4 + minor general editing. Saved new version that we should use from now on.

\usepackage[myheadings]{fullpage}
\usepackage{pmetrika}
\usepackage{pmbib}
\usepackage{color}
\usepackage{mathtools}
\usepackage{amssymb}
\usepackage{hyperref}
\usepackage{natbib}
%\usepackage[backend=biber,natbib=true, style=authoryear-comp]{biblatex}

@@

%\title{Something that is not PCADSC}
%\title{Structure comparison for multivariate data by\ldots / with application to \ldots}
\title{Exploratory data structure comparisons using Principal Component Analysis}

@@

Classical statistical methodology is aimed at analyzing data from designed experiments and historically statistical analyses have been done by researchers who knew the design and origin story of the data set well. The origin stories of data sets have changed over time and today a lot of data is accumulated without specific purpose. This is due to vast amounts of data being registered online and to a trend towards more open source research. The latter phenomenon in particular poses new challenges wrt. data quality assessment. When data are collected and made public without a specific end-point in mind, how do we ensure that differences in, say, choice measurement instruments, mode of administration, or sampling frame do not cause the data to be effectively divided into subsets that are simply not comparable?

@@

Sophisticated methods for addressing this question are available when we are willing to assume a statistical model, but when these models are taken away, a remarkable void of methods is left behind. What is needed is a procedure that compares differences in overall data structures in two (or more) subsets of a dataset without assuming neither directional nor hierarchical relationships between the variables. We propose a new method for this task, namely Principal Component Analysis-based Data Structure Comparison (PCADSC). This method employs the principal component decomposition of the data matrix performed on two subsets of a dataset in order to create intuitive visualizations of data structure differences. \hl{Mention R package.}

This manuscript is structured as follows: First, in Section \ref{sec:stateoftheart}, we present the data structure comparison problem in more detail and discuss what statistical methods are already available for solving similar challenges. Next, in Section \ref{sec:pcadscintro}, we move on to a description of the PCADSC procedure, including a brief introduction to principal component analysis (PCA) in general. In Section \ref{sec:dataexample}, we present a worked data example using the open source, online available PISA data (\hl{ref}), which is an example of a dataset where multiple data collection methods \hl{Eller måske lande?} have been employed.

@@

\section{PCADSC - description of the method}
\label{sec:pcadscintro}
\hl{Description from Anne's master's thesis. Rewrite.}

As mentioned above, the purpose of PCADSC is comparing overall data structures in two or more subsets of a dataset. But before we can get further into describing this procedure, we must first define what exactly is meant by "overall structure". One such definition is the structure of the covariance matrix of the dataset. If we assume all variables in the dataset to be jointly normal with zero means, the covariance matrix is a sufficient statistic for describing the simultaneous distribution of all the variables. This gives it a very nice interpretation as a measure of the overall structure. If we do not accept the normality assumption, pairwise correlations and variable variances are still interesting quantities that say something about the interrelations between the variables, at least in terms of linear relationships. All in all, the empirical covariance matrix is a reasonable place to start looking for differences in "overall data structures".

Though the idea sounds appealing, it is quite difficult to assess similarity of matrices, and moreover, this becomes increasingly difficult for large numbers of variables and thus high dimensional covariance matrices. There is simply too much information to consider at once.

However, by 
%clever 
use of linear algebra, we can construct a decomposition of the covariance matrix that makes it easier to gain an overview of the data. We propose a new method based on principal component analysis that seems to be able to identify differences in datasets based on intuitive, visual inspections. We refer to this method as principal component analysis-based data structure comparison (PCADSC) and we present the procedure below. But first, we give a minimal introduction to principal component analysis in general with reference to Koch (2014).

@@

Consider $n$ observations $x_1,\dotsc,x_n \in \RR^d$ of $d$ variables, let $\bar{x} = \frac{1}{n} \sum_{i=1}^n x_n$ denote their average\hl{(s?)} and let $S = \frac{1}{n-1} \sum_{i=1}^n (x_i-\bar{x}) (x_i-\bar{x})^\top \in \RR^{d \times d}$ denote the empirical covariance \hl{Is $\bar{x} \in \RR^d$ with $\bar{x} = (\bar{x_1}, ..., \bar{x_d})^T$?}. Suppose that we want to describe the observations by $q$ numbers instead of the original $d$ numbers. The associated \emph{rank-q-reconstruction error} is defined as the minimal squared error that is achievable by linear subspaces $K_q \subset \RR^d$ of dimension $q < d$, that is
\begin{equation*}
\min_{K_q} \sum_{i=1}^n \min_{z \in K_q} \lVert x_i - \bar{x} - z \rVert^2 =
\min_{K_q} \sum_{i=1}^n \lVert x_i - \bar{x} - \text{proj}_{K_q}(x_i - \bar{x}) \rVert^2.
\end{equation*}
\emph{Principal component analysis} (PCA) ensures the existence of a subspace $\hat{K}_q \subset \RR^d$ that attains this minimum, and it provides an explicit description of $\hat{K}_q$ and the rank-q-reconstruction error. Thus, let $S = U \Lambda U^\top$ be the eigenvalue decomposition of $S$. Here $\Lambda \in \RR^d$ is the diagonal matrix with the eigenvectors $\lambda_1 \ge \dotsm \ge \lambda_d \ge 0$ in the diagonal, and $U \in \RR^d$ is the orthogonal matrix with the associated eigenvectors $\eta_1,\dotsc,\eta_d \in \RR^d$ in the columns. The eigenvalues are uniquely defined, and the eigenvectors are uniquely defined up to a change of sign whenever the eigenvalues are different. If some of the eigenvalues are identical, e.g.\ $\lambda_i=\lambda_{i+1}=\dotsm=\lambda_j$, then the associated eigenvectors $\eta_i,\eta_{i+1},\dotsc,\eta_j$ are uniquely defined up to a common rotation. In practice this only happens if $n < d$, in which case the last $d-n$ eigenvalues will be zero \hl{Anne: What about covariance matrices that are not of full rank (e.g. because of colinearity)? Won't we find eigenvalue multiplicities $> 1$ here as well?}. It is a result from linear algebra that the rank-q-reconstruction error for $q < d$ is achieved for
\begin{equation*}
\hat{K}_q = \text{span}\{\eta_1,\dotsc,\eta_q\}
\end{equation*}
and equals $\sum_{j=q+1}^d \lambda_j$. The eigenvectors $\eta_j \in \RR^d$ are called \emph{loadings}, and the eigenvalues $\lambda_j \ge 0$ may be understood as \emph{variation components}. The projections $\eta_j^\top (x_i - \bar{x})$ of the observations onto the loadings are called \emph{scores}. The $j$th loading can also be found iteratively as the unit vector $u \in \RR^d$ orthogonal to $\hat{K}_{j-1}$, where the initial subspace is defined as $\hat{K}_0 = \{0\}$, that maximizes the variation of the associated scores:
\begin{align*}
\eta_j &= \argmax_{u \in \RR^d\colon u \perp \hat{K}_{j-1}} \sum_{i=1}^n \lVert u^\top (x_i - \bar{x}) \rVert^2, &
\lambda_j &= \frac{1}{n-1} \sum_{i=1}^n \lVert \eta_j^\top (x_i - \bar{x}) \rVert^2.
\end{align*}
\hl{Anne: Should we maybe assume standardized $x$s here also? I feel like it is part of the PCA algorithm and subtracting the mean in the above seems off for that reason. Also: Something is wrong with the dimensions here. $u$ is $d \times 1$, $x_i$ is $d \times 1$(?) - how do we get $\eta_j$ to also be $d \times 1$?} It is worth emphasizing that the greedy approach of successively adding the next direction $\eta_j$ explaining most of the remaining variation, also gives the sequence $\hat{K}_q = \hat{K}_{q-1} \oplus \text{span} \{\eta_q\}$ of subspaces minimizing the rank-q-reconstruction error. This strong interpretation of PCA, which is often overlooked in the literature, means that the sequence of loadings $\eta_j$ and their associated variation components $\lambda_j$ yield a simultaneous description of the structure of the data set for all approximating dimensions $q$. This implies that the loadings and variation components can be used to investigate the structure of the data set without the need to decide on an approximating dimension $q$.

@@

In this paper we propose three diagnostic plots for comparing the loading patterns in two datasets $x_{11},\dotsc,x_{1 n_1} \in \RR^d$ and $x_{21},\dotsc,x_{2 n_2} \in \RR^d$ in the same $d$ variables. The construction of these plots, which we call \emph{CumEigenPlot}, \emph{HairPlot} and \emph{PancakePlot}, proceeds in the following steps:
\begin{enumerate}
\item Standardized the observations $x_{i1},\dotsc,x_{i n_i}$ to have zero mean and unit standard deviation separately for each of the $d$ variables within each datasets. Let $\tilde{x}_{ij} \in \RR^d$ for $i=1,2$ and $j=1,\dotsc,n_i$ be the standardized observations, and form the principal component analysis
\begin{equation*}
S_i = \frac{1}{n_i-1} \sum_{j=1}^{n_i} \tilde{x}_{ij} \tilde{x}_{ij}^\top = \sum_{k=1}^d \lambda_{ik} \eta_{ik} \eta_{ik}^\top \quad \text{for $i=1,2$.} 
\end{equation*}
\item Combine the standardized datasets into a joint dataset $\tilde{x}_1,\dotsc,\tilde{x}_n \in \RR^d$ with $n=n_1+n_2$ observations, standardize this dataset, and form the principal component analysis
\begin{equation*}
S = \frac{1}{n-2} \sum_{j=1}^n \tilde{x}_j \tilde{x}_j^\top = \sum_{k=1}^d \lambda_k \eta_k \eta_k^\top.
\end{equation*}
\item Make the diagnostic plot of interest. Details of the plots will be described below.
\item To investigate the sample variation randomly reallocate the combined and standardized dataset formed in step 2 to two datasets of $n_1$ and $n_2$ observations, respectively, and redo step 1 and 3. This is done several times, e.g.\ 10000 times, and compared to the plot for the original allocation.
\end{enumerate}

Before describing the details of the  \emph{CumEigenPlot}, \emph{HairPlot} and \emph{PancakePlot} we make some general remarks. First, it should be noted that PCA is sensitive to scaling, as the procedure deconstructs the covariance matrix in components according to the most explained variance. This implies that if a variable has a very large sample variance (possibly because of its scale), this variable will be always be deemed highly influential, no matter the structure of the data. Therefore, the variables should always be scaled prior to performing PCA. Note that the covariance matrix for the standardized variables is the same as the correlation matrix for the original variables, so this simply corresponds to performing data structural comparisons of the correlation matrices rather than the covariance matrices. \hl{Do we still need more about this topic? Then we can add something like:
\begin{itemize}
\item Following the greedy interpretation of PCA, we know that the first eigenvector is chosen as
$$\eta_1 = \argmax_{u \in \RR^d} \sum_{i=1}^n \lVert u^\top (x_i - \bar{x}) \rVert^2  = \text{argmax}_{u \in \RR^d} \tilde{\V}(u^T X^T) $$
\item Now, imagine that $\V(x_1) >> \V(x_j)$ for all $j \in 2, ..., d$. Note that $(1, 0, ..., 0) X^T = x_i$. Then, clearly a vector $u$ that only picks out the first variable from the data matrix $X$, namely $u^T = (1, 0, ..., 0)$ will allow us to only look at the variance of $x_1$, which is very large. This has nothing to do with the interplay between variables, as the phenomenon will disappear if we scale $x_1$
\item A more formal explanation could come from this kind of intuition, if needed.
\end{itemize}
}
The standardization makes the variables comparable on the same scale, i.e.\ units of standard deviation, and it implies that the diagonal elements of $S$, $S_1$, and $S_2$ all equals 1. This also holds for the reallocated datasets, and in particular we have $\sum_{k=1}^d \lambda_k = \sum_{k=1}^d \lambda_{1k} = \sum_{k=1}^d \lambda_{2k} =  d$. The random reallocations in step 4 provide resamples under the null hypothesis that the two sample populations have the same structure in the $d$ variables, except that the two populations are allowed to have the own standard deviations for each of the variables. This exception is induced by the initial step 1, which performs a standardization within the two datasets before they are combined in step 2. These resamples can be used for a visualization of the sample variation, and if a test statistic can be defined in step 3, then a $p$-value for the null hypothesis may be computed. Such $p$-values are know as permutation tests for Goodness-of-Fit (\hl{reference?}).

@@

\subsubsection{Description of the \emph{CumEigenPlot}}
 This plot compares the variation components. 
%If these are the same in the two sample populations, then the best estimate for the variation components are $\lambda_1 \ge \dotsm \ge \lambda_d \ge 0$ found in the combined dataset. And we would expect $\lambda_{i1} \ge \dotsm \ge \lambda_{id} \ge 0$ for $i=1,2$ to be alike excepts sample variation. 
In order to investigate whether the same proportion of the total variation can be described by the same number of principal components in the two datasets we plot a piecewise linear curve connecting the points
\begin{align*}
(0,0), &&
(\lambda_1,\lambda_{11}-\lambda_{12}), &&
(\lambda_1 + \lambda_2,\lambda_{11}+\lambda_{12}-\lambda_{21}-\lambda_{22}), &&
\ldots, &&
\bigg( \sum_{j=1}^d \lambda_j, \sum_{j=1}^d \lambda_{1j} - \sum_{j=1}^d \lambda_{2j} \bigg).
\end{align*}
This may be seen as a cumulated Bland-Altman plot for the variation components (\hl{reference to cumulated residuals and to Bland-Altman}). Note that due to the standardization the last point will be equal to $(d,0)$. Thus, this curve will begin and end at the x-axis. And the larger excursions it makes away from the x-axis the less alike the cumulated variation components for the two datasets are. \hl{Is this an informative x-axis? Should we maybe consider scaling it?}

@@

\subsubsection{Description of the \emph{HairPlot}} 
This plot simultaneously compares the variation components and the loadings. Let $\lambda_{\max} = \max\{ \lambda_{11}, \lambda_{21} \}$ be the largest variation component for the two datasets. The empirical correlation matrix $S_1$ for the first dataset has the following orthogonal decomposition in the coordinate system of the second dataset
%\begin{equation*}
%S_1 = \sum_{k=1}^d \lambda_{1k} \eta_{1k} \eta_{1k}^\top
%= \lambda_{\max} \sum_{k=1}^d
%\Bigg( \sum_{j=1}^d \sqrt{\frac{\lambda_{1k}}{\lambda_{\max}}} (\eta_{1k} \eta_{2j}^\top) \eta_{2j} \Bigg)
%\Bigg( \sum_{j=1}^d \sqrt{\frac{\lambda_{1k}}{\lambda_{\max}}} (\eta_{1k} \eta_{2j}^\top) \eta_{2j} \Bigg)^\top,
%\end{equation*}
\begin{equation*}
S_1 = \sum_{k=1}^d \lambda_{1k} \eta_{1k} \eta_{1k}^\top
= \lambda_{\max} \sum_{k=1}^d
\Bigg( \sum_{j=1}^d \sqrt{\frac{\lambda_{1k}}{\lambda_{\max}}} \eta_{2j} (\eta_{2j}^\top \eta_{1k}) \Bigg)
\Bigg( \sum_{j=1}^d \sqrt{\frac{\lambda_{1k}}{\lambda_{\max}}} \eta_{2j} (\eta_{2j}^\top \eta_{1k}) \Bigg)^\top,
\end{equation*}
and we have a similar decomposition of $S_2$ in the coordinate system of the first dataset \hl{Anne: it was not completely clear to me what this meant the first xx times I read it - expand?}. We propose to visualize these two decompositions in a $d \times d$ grid display. In the $j$th row and $k$th column of this display we plot two arrows based at the lower left corner of the grid cell. The first arrow has length $\mu_{jk}$ and angle $\theta_{jk}/2$ anticlockwise from the diagonal, and the second arrow has length $\nu_{jk}$ and angle $\theta_{jk}/2$ clockwise from the diagonal. To facilitate the following description we will refer to the arrows drawn anticlockwise as the blue arrows, and the arrows drawn clockwise as the red arrows. The lengths $\mu_{jk}$ and $\nu_{jk}$ and the angle $\theta_{kj}$ are given by
\begin{align*}
\mu_{jk} &= \sqrt{\frac{\lambda_{1k}}{\lambda_{\max}}} \lvert \eta_{1k}^\top \eta_{2j} \rvert, &
\nu_{jk} &= \sqrt{\frac{\lambda_{2j}}{\lambda_{\max}}} \lvert \eta_{2j}^\top \eta_{1k} \rvert, &
\theta_{jk} &= \acos(\lvert \eta_{1k}^\top \eta_{2j} \rvert).
\end{align*}
\hl{Maybe emphasize inner product/angle-interpretation here?}
The absolute value of the projection $\eta_{1k} \eta_{2j}^\top$ is inserted due to the indeterminacy of the direction of loading vectors. This indeterminacy implies that the angle between loadings from the two datasets always can be chosen to be in the interval $[0,\pi/2]$, and hence the decomposition of $S_1$ and $S_2$ can be visualized in a joint plot by dividing the angles by two and using anticlockwise and clockwise shifts from the diagonal. Furthermore, the scaling of the lengths by $\lambda_{\max}$ is made so that the longest arrow has at most unit length.

In the \emph{HairPlot} the blue arrows in the $k$th column of the grid display visualize the decomposition of the $k$th principal component for the first dataset in the coordinate system of the second dataset. Similarly, the red arrows in the $j$th row visualize the decomposition of the $j$th principal components for the second dataset in the coordinate system of the first dataset. \hl{Can we expand on this in a non-technical way?} If the structures of the two datasets are identical, then we will have coinciding blue and red arrows along the diagonal in the grid display, and nothing else as arrows in the off-diagonal cells would have zero length. Differences in the variation components are visualized as differences in the lengths of the blue and the red arrows, also in the diagonal. And loadings in other directions than the corresponding loading from the other dataset are visualized as angle separation of the blue and the red arrows in the diagonal cells as well as arrows of non vanishing length in the off-diagonal cells. The name \emph{HairPlot} arise by seeing the arrows as \emph{hairs}. Using this interpretation we may say that a more ``hairy'' plot indicates less agreement between the two loading patterns.

>>>>

\usepackage[myheadings]{fullpage}
\usepackage{pmetrika}
\usepackage{pmbib}
\usepackage{color}
\usepackage{mathtools}
\usepackage{amssymb}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{multirow}
%\usepackage[backend=biber,natbib=true, style=authoryear-comp]{biblatex}

@@

%\title{Something that is not PCADSC}
%\title{Structure comparison for multivariate data by\ldots / with application to \ldots}
\title{Exploratory data structure comparisons: Three new tools based on Principal Component Analysis}

@@

Classical statistical methodology is aimed at analyzing data from designed experiments and historically, statistical analyses have been done by researchers who knew the design and origin story of the data set well. However, the origin stories of data sets have changed over time and today a lot of data is accumulated without a specific purpose in mind. This is due to vast amounts of data being registered online and to a trend towards more open source research. The latter phenomenon in particular poses new challenges wrt. data quality assessment. When data are collected and made public without a specific end-point in mind, how do we ensure that differences in, say, choice measurement instruments, mode of administration, or sampling frame do not cause the data to be effectively divided into subsets that are simply not comparable?

@@

Sophisticated methods for addressing this question are available when we are willing to assume a statistical model, but when these models are taken away, a remarkable void of methods is left behind. What is needed is a procedure that compares differences in overall data structures in two (or more) subsets of a dataset without assuming neither directional nor hierarchical relationships between the variables. We propose a suite of three new tools for this task, which we will refer to collectively as Principal Component Analysis-based Data Structure Comparisons (PCADSC). These methods employ the principal component decomposition of the empirical covariance matrix performed on two subsets of a dataset in order to create intuitive visualizations of data structure differences. Software for using these tools is available in an \texttt{R}-package. \hl{how to refer to our package?}

This manuscript is structured as follows: First, in Section \ref{sec:stateoftheart}, we present the data structure comparison problem in more detail and discuss what statistical methods are already available for solving similar challenges. Next, in Section \ref{sec:pcadscintro}, we move on to a description of the PCADSC procedures, including a brief introduction to principal component analysis (PCA) in general. In Section \ref{sec:dataexample}, we present a worked data example using open source, online available data on psychological well-being in three European countries. More specifically, we compare data from Denmark, which has been repeatedly been rewarded "happiest country in the world", with data from Bulgaria and Sweden, respectively, to investigate whether or not psychological well-being - and thus happiness - is really a concept that is universal beyond country borders, or if such rankings of happiness are inherently meaningless. At last, we discuss limitations and merits of the PCADSC tools in Section \ref{sec:discussion}.

@@

\section{PCA-based tools for data structure comparisons}
\label{sec:pcadscintro}
As mentioned above, the purpose of PCADSC is comparing overall data structures in two or more subsets of a dataset. But before we can get further into describing the PCADSC tools, we must first define the exact meaning of \textit{overall structures} in this context. One such definition is the structure of the covariance matrix of the dataset. If we assume all variables in the dataset to be jointly normal with zero means, the covariance matrix is a sufficient statistic for describing the simultaneous distribution of all the variables. This gives it a very nice interpretation as a measure of the overall structure. If we do not accept the normality assumption, pairwise correlations and marginal variable variances are still interesting quantities that say something about the interrelations between the variables in the data. All in all, the empirical covariance matrix is a reasonable place to start looking for differences in \textit{overall data structures}.

Though the idea sounds appealing at first, it is quite difficult to assess similarity of matrices, and moreover, it becomes increasingly difficult for large numbers of variables and thus high dimensional covariance matrices. There is simply too much information to consider at once. However, by use of linear algebra, we can decompose and recompose the covariance matrix, such that the distinct dimensions of information withheld in it are clearly separated and ordered according to their relative importance. In particular, we find a new representation of the covariance matrix that makes it to gain an overview of the most interesting aspects of the data. On such decomposition strategy is using \textit{principal component analysis} (PCA), which uses eigenvalue decomposition in order to obtain a new representation of the covariance matrix.

@@

Consider $n$ observations $x_1,\dotsc,x_n \in \RR^d$ of $d$ variables, let $\bar{x} = \frac{1}{n} \sum_{i=1}^n x_n$ denote their average\hl{(s?)} and let $S = \frac{1}{n-1} \sum_{i=1}^n (x_i-\bar{x}) (x_i-\bar{x})^\top \in \RR^{d \times d}$ denote the empirical covariance \hl{Is $\bar{x} \in \RR^d$ with $\bar{x} = (\bar{x_1}, ..., \bar{x_d})^T$?}. Suppose that we want to describe the observations by $q$ numbers instead of the original $d$ numbers. The associated \emph{rank-q-reconstruction error} is defined as the minimal squared error that is achievable by linear subspaces $K_q \subset \RR^d$ of dimension $q < d$, that is
\begin{equation*}
\min_{K_q} \sum_{i=1}^n \min_{z \in K_q} \lVert x_i - \bar{x} - z \rVert^2 =
\min_{K_q} \sum_{i=1}^n \lVert x_i - \bar{x} - \text{proj}_{K_q}(x_i - \bar{x}) \rVert^2.
\end{equation*}
\emph{Principal component analysis} (PCA) ensures the existence of a subspace $\hat{K}_q \subset \RR^d$ that attains this minimum, and it provides an explicit description of $\hat{K}_q$ and the rank-q-reconstruction error. Thus, let $S = U \Lambda U^\top$ be the eigenvalue decomposition of $S$. Here $\Lambda \in \RR^d$ is the diagonal matrix with the eigenvectors $\lambda_1 \ge \dotsm \ge \lambda_d \ge 0$ in the diagonal, and $U \in \RR^d$ is the orthogonal matrix with the associated eigenvectors $\eta_1,\dotsc,\eta_d \in \RR^d$ in the columns. The eigenvalues are uniquely defined, and the eigenvectors are uniquely defined up to a change of sign whenever the eigenvalues are different. If some of the eigenvalues are identical, e.g.\ $\lambda_i=\lambda_{i+1}=\dotsm=\lambda_j$, then the associated eigenvectors $\eta_i,\eta_{i+1},\dotsc,\eta_j$ are uniquely defined up to a common rotation. In practice this only happens if $n < d$, in which case the last $d-n$ eigenvalues will be zero \hl{Anne: What about covariance matrices that are not of full rank (e.g. because of colinearity)? Won't we find eigenvalue multiplicities $> 1$ here as well?}. It is a result from linear algebra that the rank-q-reconstruction error for $q < d$ is achieved for
\begin{equation*}
\hat{K}_q = \text{span}\{\eta_1,\dotsc,\eta_q\}
\end{equation*}
and equals $\sum_{j=q+1}^d \lambda_j$. The eigenvectors $\eta_j \in \RR^d$ are called \emph{loadings}, and the eigenvalues $\lambda_j \ge 0$ may be understood as \emph{variation components}. The projections $\eta_j^\top (x_i - \bar{x})$ of the observations onto the loadings are called \emph{scores}. The $j$th loading can also be found iteratively as the unit vector $u \in \RR^d$ orthogonal to $\hat{K}_{j-1}$, where the initial subspace is defined as $\hat{K}_0 = \{0\}$, that maximizes the variation of the associated scores:
\begin{align*}
\eta_j &= \argmax_{u \in \RR^d\colon u \perp \hat{K}_{j-1}} \sum_{i=1}^n \lVert u^\top (x_i - \bar{x}) \rVert^2, &
\lambda_j &= \frac{1}{n-1} \sum_{i=1}^n \lVert \eta_j^\top (x_i - \bar{x}) \rVert^2.
\end{align*}
\hl{Anne: Should we maybe assume standardized $x$s here also? I feel like it is part of the PCA algorithm and subtracting the mean in the above seems off for that reason. Also: Something is wrong with the dimensions here. $u$ is $d \times 1$, $x_i$ is $d \times 1$(?) - how do we get $\eta_j$ to also be $d \times 1$?} It is worth emphasizing that the greedy approach of successively adding the next direction $\eta_j$ explaining most of the remaining variation, also gives the sequence $\hat{K}_q = \hat{K}_{q-1} \oplus \text{span} \{\eta_q\}$ of subspaces minimizing the rank-q-reconstruction error. This strong interpretation of PCA, which is often overlooked in the literature, means that the sequence of loadings $\eta_j$ and their associated variation components $\lambda_j$ yield a simultaneous description of the structure of the data set for all approximating dimensions $q$. This implies that the loadings and variation components can be used to investigate the structure of the data set without the need to decide on an approximating dimension $q$.

@@

In this paper we propose three diagnostic plots for comparing the loading patterns in two datasets $x_{11},\dotsc,x_{1 n_1} \in \RR^d$ and $x_{21},\dotsc,x_{2 n_2} \in \RR^d$ in the same $d$ variables. The construction of these plots, which we call \textit{CE plot}, \textit{angle plot} and \textit{chroma plot}, proceeds in the following steps:
\begin{enumerate}
\item Standardized the observations $x_{i1},\dotsc,x_{i n_i}$ to have zero mean and unit standard deviation separately for each of the $d$ variables within each datasets. Let $\tilde{x}_{ij} \in \RR^d$ for $i=1,2$ and $j=1,\dotsc,n_i$ be the standardized observations, and form the principal component analysis
\begin{equation*}
S_i = \frac{1}{n_i-1} \sum_{j=1}^{n_i} \tilde{x}_{ij} \tilde{x}_{ij}^\top = \sum_{k=1}^d \lambda_{ik} \eta_{ik} \eta_{ik}^\top \quad \text{for $i=1,2$.} 
\end{equation*}
\item Combine the standardized datasets into a joint dataset $\tilde{x}_1,\dotsc,\tilde{x}_n \in \RR^d$ with $n=n_1+n_2$ observations, standardize this dataset, and form the principal component analysis
\begin{equation*}
S = \frac{1}{n-2} \sum_{j=1}^n \tilde{x}_j \tilde{x}_j^\top = \sum_{k=1}^d \lambda_k \eta_k \eta_k^\top.
\end{equation*}
\item Make the diagnostic plot of interest. Details of the plots will be described below.
\item To investigate the sample variation randomly reallocate the combined and standardized dataset formed in step 2 to two datasets of $n_1$ and $n_2$ observations, respectively, and redo step 1 and 3. This is done several times, e.g.\ 10000 times, and compared to the plot for the original allocation.
\end{enumerate}

Before describing the details of the  \emph{CE plot}, the \emph{angle plot} and \emph{chroma plot}, we make some general remarks. First, it should be noted that PCA is sensitive to scaling, as the procedure deconstructs the covariance matrix in components according to the most explained variance. This implies that if a variable has a very large sample variance (possibly because of its scale), this variable will be always be deemed highly influential, no matter the structure of the data. Therefore, the variables should always be scaled prior to performing PCA. Note that the covariance matrix for the standardized variables is the same as the correlation matrix for the original variables, so this simply corresponds to performing data structural comparisons of the correlation matrices rather than the covariance matrices. \hl{Do we still need more about this topic? Then we can add something like:
\begin{itemize}
\item Following the greedy interpretation of PCA, we know that the first eigenvector is chosen as
$$\eta_1 = \argmax_{u \in \RR^d} \sum_{i=1}^n \lVert u^\top (x_i - \bar{x}) \rVert^2  = \text{argmax}_{u \in \RR^d} \tilde{\V}(u^T X^T) $$
\item Now, imagine that $\V(x_1) >> \V(x_j)$ for all $j \in 2, ..., d$. Note that $(1, 0, ..., 0) X^T = x_i$. Then, clearly a vector $u$ that only picks out the first variable from the data matrix $X$, namely $u^T = (1, 0, ..., 0)$ will allow us to only look at the variance of $x_1$, which is very large. This has nothing to do with the interplay between variables, as the phenomenon will disappear if we scale $x_1$
\item A more formal explanation could come from this kind of intuition, if needed.
\end{itemize}
}
The standardization makes the variables comparable on the same scale, i.e.\ units of standard deviation, and it implies that the diagonal elements of $S$, $S_1$, and $S_2$ all equals 1. This also holds for the reallocated datasets, and in particular we have $\sum_{k=1}^d \lambda_k = \sum_{k=1}^d \lambda_{1k} = \sum_{k=1}^d \lambda_{2k} =  d$. The random reallocations in step 4 provide resamples under the null hypothesis that the two sample populations have the same structure in the $d$ variables, except that the two populations are allowed to have the own standard deviations for each of the variables. This exception is induced by the initial step 1, which performs a standardization within the two datasets before they are combined in step 2. These resamples can be used for a visualization of the sample variation, and if a test statistic can be defined in step 3, then a $p$-value for the null hypothesis may be computed. Such $p$-values are know as permutation tests for Goodness-of-Fit (\hl{reference?}).

@@

\subsubsection{The cumulative eigenvalue plot}
 The cumulative eigenvalue (CE) plot compares the variation components, i.e. the eigenvalues of the covariance matrix. 
%If these are the same in the two sample populations, then the best estimate for the variation components are $\lambda_1 \ge \dotsm \ge \lambda_d \ge 0$ found in the combined dataset. And we would expect $\lambda_{i1} \ge \dotsm \ge \lambda_{id} \ge 0$ for $i=1,2$ to be alike excepts sample variation. 
In order to investigate whether the same proportion of the total variation can be described by the same number of principal components in the two datasets we plot a piecewise linear curve connecting the points
\begin{align*}
(0,0), &&
(\lambda_1,\lambda_{11}-\lambda_{12}), &&
(\lambda_1 + \lambda_2,\lambda_{11}+\lambda_{12}-\lambda_{21}-\lambda_{22}), &&
\ldots, &&
\bigg( \sum_{j=1}^d \lambda_j, \sum_{j=1}^d \lambda_{1j} - \sum_{j=1}^d \lambda_{2j} \bigg).
\end{align*}
This may be seen as a cumulative Bland-Altman plot for the variation components (\hl{reference to cumulative residuals and to Bland-Altman}). Note that due to the standardization the last point will be equal to $(d,0)$. Thus, this curve will begin and end at the x-axis. And the larger excursions it makes away from the x-axis the less alike the cumulative variation components for the two datasets are. \hl{Is this an informative x-axis? Should we maybe consider scaling it?}

@@

\subsubsection{The angle plot} 
This plot simultaneously compares the variation components and the loadings. Let $\lambda_{\max} = \max\{ \lambda_{11}, \lambda_{21} \}$ be the largest variation component for the two datasets. The empirical correlation matrix $S_1$ for the first dataset has the following orthogonal decomposition in the coordinate system of the second dataset
%\begin{equation*}
%S_1 = \sum_{k=1}^d \lambda_{1k} \eta_{1k} \eta_{1k}^\top
%= \lambda_{\max} \sum_{k=1}^d
%\Bigg( \sum_{j=1}^d \sqrt{\frac{\lambda_{1k}}{\lambda_{\max}}} (\eta_{1k} \eta_{2j}^\top) \eta_{2j} \Bigg)
%\Bigg( \sum_{j=1}^d \sqrt{\frac{\lambda_{1k}}{\lambda_{\max}}} (\eta_{1k} \eta_{2j}^\top) \eta_{2j} \Bigg)^\top,
%\end{equation*}
\begin{equation*}
S_1 = \sum_{k=1}^d \lambda_{1k} \eta_{1k} \eta_{1k}^\top
= \lambda_{\max} \sum_{k=1}^d
\Bigg( \sum_{j=1}^d \sqrt{\frac{\lambda_{1k}}{\lambda_{\max}}} \eta_{2j} (\eta_{2j}^\top \eta_{1k}) \Bigg)
\Bigg( \sum_{j=1}^d \sqrt{\frac{\lambda_{1k}}{\lambda_{\max}}} \eta_{2j} (\eta_{2j}^\top \eta_{1k}) \Bigg)^\top,
\end{equation*}
and we have a similar decomposition of $S_2$ in the coordinate system of the first dataset \hl{Anne: it was not completely clear to me what this meant the first xx times I read it - expand?}. We propose to visualize these two decompositions in a $d \times d$ grid display. In the $j$th row and $k$th column of this display we plot two arrows based at the lower left corner of the grid cell. The first arrow has length $\mu_{jk}$ and angle $\theta_{jk}/2$ anticlockwise from the diagonal, and the second arrow has length $\nu_{jk}$ and angle $\theta_{jk}/2$ clockwise from the diagonal. To facilitate the following description we will refer to the arrows drawn anticlockwise as the blue arrows, and the arrows drawn clockwise as the red arrows. The lengths $\mu_{jk}$ and $\nu_{jk}$ and the angle $\theta_{kj}$ are given by
\begin{align*}
\mu_{jk} &= \sqrt{\frac{\lambda_{1k}}{\lambda_{\max}}} \lvert \eta_{1k}^\top \eta_{2j} \rvert, &
\nu_{jk} &= \sqrt{\frac{\lambda_{2j}}{\lambda_{\max}}} \lvert \eta_{2j}^\top \eta_{1k} \rvert, &
\theta_{jk} &= \acos(\lvert \eta_{1k}^\top \eta_{2j} \rvert).
\end{align*}
\hl{Maybe emphasize inner product/angle-interpretation here?}
The absolute value of the projection $\eta_{1k} \eta_{2j}^\top$ is inserted due to the indeterminacy of the direction of loading vectors. This indeterminacy implies that the angle between loadings from the two datasets always can be chosen to be in the interval $[0,\pi/2]$, and hence the decomposition of $S_1$ and $S_2$ can be visualized in a joint plot by dividing the angles by two and using anticlockwise and clockwise shifts from the diagonal. Furthermore, the scaling of the lengths by $\lambda_{\max}$ is made so that the longest arrow has at most unit length.

In the \emph{angle plot}, the blue arrows in the $k$th column of the grid display visualize the decomposition of the $k$th principal component for the first dataset in the coordinate system of the second dataset. Similarly, the red arrows in the $j$th row visualize the decomposition of the $j$th principal components for the second dataset in the coordinate system of the first dataset. \hl{Can we expand on this in a non-technical way?} If the structures of the two datasets are identical, then we will have coinciding blue and red arrows along the diagonal in the grid display, and nothing else as arrows in the off-diagonal cells would have zero length. Differences in the variation components are visualized as differences in the lengths of the blue and the red arrows, also in the diagonal. And loadings in other directions than the corresponding loading from the other dataset are visualized as angle separation of the blue and the red arrows in the diagonal cells as well as arrows of non vanishing length in the off-diagonal cells.
